Quantitative Benchmarks for Theory Exploration

### 2017-04-25 PhDSymposium2017/abstract/abstract.tex
Submitted version for PhDSymp17

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Exploring Software with Symbolic and Statistical Algorithms},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}
\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{"#1"}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

\title{Exploring Software with Symbolic and Statistical Algorithms}
  \author{
          Chris Warburton \\
      University of Dundee \\
      \texttt{cmwarburton@dundee.ac.uk}
    }
\date{}

\begin{document}
\maketitle
\begin{abstract}
We aim to aid programmers in understanding and maintaining software
by increasing the capability of \emph{theory exploration} systems for
discovering novel, unexpected properties of programs and libraries.
Existing theory exploration systems find equivalent program fragments via brute-force
search, which doesn't scale well to realistic codebases. We narrow down
the scope of each exploration run using statistics derived from the
semantic similarity of the source code being explored, using a process
called \emph{recurrent clustering}.
\end{abstract}

@@

As computers become ever more important in research, industry and
society, so does the ability to program them effectively, economically
and \emph{correctly}. Whilst tools exist to help programmers understand
and manipulate software, ranging from linters \cite{Johnson78lint} through to formal
verification systems \cite{boyer1983proof}, these are often limited in scope (e.g. to syntax rather than semantics) or require precise guidance from the user. One
reason is that automating the search for patches \cite{Forrest.Nguyen.Weimer.ea:2009}, refactorings,
optimisations \cite{phothilimthana2016scaling}, proofs \cite{rosen2012proving}, and other routine programming tasks is
prohibitively expensive due to the size of the search space.

We investigate the use of machine learning techniques to restrict such
search spaces more intelligently than uninformed local search,
specifically the use of \emph{clustering} to aid \emph{theory
exploration} (TE) in discovering program equivalences.

@@

\subsection{Theory Exploration}\label{theory-exploration}

Our work builds on the existing TE system
\textsc{QuickSpec}\cite{QuickSpec}, which finds equivalent expressions
in the Haskell programming language \cite{marlow2010haskell}, built from a user-provided
\emph{signature} of constants and variables. For example, given a
signature of the list-manipulating functions \texttt{reverse},
\texttt{length} and \texttt{append}, plus two list variables \texttt{x}
and \texttt{y}, \textsc{QuickSpec} will discover that
\texttt{reverse\ (reverse\ x)} is equivalent to \texttt{x}; that
\texttt{length\ (reverse\ x)} is equivalent to \texttt{length\ x}; that
\texttt{length\ (append\ x\ y)} is equivalent to
\texttt{length\ (append\ y\ x)}; and so on. These are discovered by
enumerating all well-typed combinations of the signature's constants and
variables, then randomly instantiating the variables \cite{claessen2011quickcheck} and comparing the
resulting closed terms. Any expressions which remain indistiguishable
after hundreds of instantiations are conjectured to be equivalent and,
after removing redundancies, are produced as output. These conjectures
can be sent through a theorem prover \cite{claessen2013automating} and used to inform tests,
optimisations, refactoring, verification, or simply to help the
programmer learn more about the code.

Due to its exponential complexity, this enumerating procedure is limited
to producing small expressions from signatures with only a few elements.
Such small signatures give little chance for serendipitous discoveries,
which undermines the algorithm's potential to present programmers with
new, unexpected information.

\subsection{Recurrent Clustering}\label{recurrent-clustering}

To reduce the user's need to cherry-pick signatures, our approach
accepts a large signature (e.g. a complete Haskell package) and uses
statistical machine learning methods to select smaller signatures
automatically, using a \emph{recurrent clustering} method inspired by
those of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and \textsc{ACL2(ml)} \cite{Heras.Komendantskaya.Johansson.ea:2013}.

Recurrent clustering attempts to cluster expressions based on their
Abstract Syntax Trees (ASTs), an example of which is shown in Figure \ref{fig:astexample}. First the recursively-structured ASTs are
transformed into vectors of fixed length, by truncating the tree
structure and listing the node labels in breadth-first order.

These labels are then turned into \emph{features} (real numbers), using the function $\phi$ which replaces keywords with fixed values and local variables with context-dependent values (known as \emph{de Bruijn indices} \cite{de1972lambda}). To replace global variables, $\phi$ first performs another round of clustering (hence the name
``recurrent''), without including the current expression; each global
variable is replaced by its cluster index found by this ``inner'' round
of clustering.

\begin{figure}
    \begin{small}
      \begin{tikzpicture}[sibling distance=0pt]
        \tikzset{sibling distance=0pt}
        \tikzset{level distance=20pt}
        \Tree[ .$\CLam$
                  $\id{a}$
                  [ .$\CCase$
                       [ .$\CVar$
                            [ .$\CLocal$
                                 $\id{a}$ ]]
                       $\id{b}$
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CVar$
                                 $\CConstructor$ ]]
                       [ .$\CAlt$
                            $\CDataAlt$
                            [ .$\CApp$
                                 [ .$\CVar$
                                      [ .$\CGlobal$
                                           $\id{even}$ ]]
                                 [ .$\CVar$
                                      [ .$\CLocal$
                                           $\id{n}$ ]]]
                            $\id{n}$ ]]]
      \end{tikzpicture}
    \end{small}
    \caption{AST for the \texttt{odd} function. The variable names \texttt{a}, \texttt{b}, etc. are chosen arbitrarily.}
    \label{fig:astexample}
\end{figure}

These recursive calls to the clustering algorithm (we use a standard
\emph{k-means} implementation) allow expressions to be compared in a
principled way: the similarity of expressions depends on the similarity of the expressions they reference, and so on recursively. In practice, this recursive
algorithm can be implemented in an iterative way, by accumulating the set of expressions to cluster in topological order of their dependency graph.

Figure \ref{fig:astexample} shows the AST for the following Haskell function \texttt{odd}, which determines if a Peano numeral is odd:

\begin{lstlisting}[language=Haskell]
odd  Z    = False
odd (S n) = even n

even  Z    = True
even (S n) = odd n
\end{lstlisting}

The \texttt{odd} function references four global variables: \texttt{Z}, \texttt{S}, \texttt{False} and \texttt{even}. Our algorithm doesn't yet distinguish between data constructors, so the only global reference we replace is \texttt{even}. Likewise, the definition of \texttt{even} references \texttt{Z}, \texttt{S}, \texttt{True} and \texttt{odd}. For mutual recursion like this, there is no valid topological order, so we use a sentinel value as the feature.

These feature vectors are sent through a k-means clustering algorithm, using the \textsc{Weka} machine learning library. For example, the feature vector for \texttt{odd} (padded to 6 labels for each level of the tree) will begin:

\begin{equation*}
  (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \dots
\end{equation*}

\section{Implementation}\label{implementation}

We obtain ASTs from Haskell projects using a bespoke plugin for the GHC
compiler. From these ASTs, we extract type information which is used to append variables to the signature; and dependency information, which is used to topologically sort the clustering rounds. We then invoke our iterative algorithm, which alternates between feature extraction and clustering until all elements of the signature have been clustered. Each cluster is converted into a \textsc{QuickSpec} signature, by extracting those functions which a) have an argument type which we can randomly generate and b) have an output type which we can compare. \textsc{QuickSpec} is invoked on each signature, and the resulting sets of conjectures are combined.

\section{Results}\label{results}

We have developed a machine learning approach for analysing Haskell code, including a
bespoke feature extraction method and a full implementation pipeline for turning Haskell packages into sets of equations.

\section{Difficulties and Future Work}\label{future-work}

The most difficult aspect of performing this exploration was obtaining real code in a usable format, due to the myriad edge-cases encountered. We make extensive use of Haskell's existing infrastructure, including the \textsc{GHC} compiler, the \textsc{cabal} build system, the \textsc{Hackage} code repository and the \textsc{Nix} package manager. Unfortunately, some of these systems are rather monolithic, which makes it difficult to invoke particular algorithms, such as \textsc{GHC}'s type class resolution, on their own. Whilst we have worked around these issues, e.g. using meta-programming, this introduces unnecessary latency, fragility and complexity.

Our investigation of recurrent clustering and theory exploration only scratches the surface of combining symbolic and statistical AI algorithms. As well as benchmarking our approach against other techniques, there are many similar opportunities to be explored, where the reasoning power of symbolic algorithms can be guided and supervised by statistical, data-driven processes.

\bibliographystyle{plain}
\bibliography{../Bibtex}

>>>>

\documentclass[]{default}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{amssymb,amsmath}
\usepackage{paralist}
\usepackage{tikz-qtree}
\usepackage{enumitem}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum0\ifxetex{1}\fi\ifluatex{1}\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex{
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  }\fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex%
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Quantitative Benchmarks for Theory Exploration},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

\newcommand{\name}[1]{\mathrm{#1}}

\newcommand{\feature}[1]{\phi(#1)}
\newcommand{\id}[1]{\texttt{``#1''}}
\newcommand{\CVar}{\texttt{Var}}
\newcommand{\CLit}{\texttt{Lit}}
\newcommand{\CApp}{\texttt{App}}
\newcommand{\CLam}{\texttt{Lam}}
\newcommand{\CLet}{\texttt{Let}}
\newcommand{\CCase}{\texttt{Case}}
\newcommand{\CType}{\texttt{Type}}
\newcommand{\CLocal}{\texttt{Local}}
\newcommand{\CGlobal}{\texttt{Global}}
\newcommand{\CConstructor}{\texttt{Constructor}}
\newcommand{\CLitNum}{\texttt{LitNum}}
\newcommand{\CLitStr}{\texttt{LitStr}}
\newcommand{\CAlt}{\texttt{Alt}}
\newcommand{\CDataAlt}{\texttt{DataAlt}}
\newcommand{\CLitAlt}{\texttt{LitAlt}}
\newcommand{\CDefault}{\texttt{Default}}
\newcommand{\CNonRec}{\texttt{NonRec}}
\newcommand{\CRec}{\texttt{Rec}}
\newcommand{\CBind}{\texttt{Bind}}

\newlist{inlinelist}{enumerate*}{1}
\setlist*[inlinelist,1]{%
  label= (\roman*),
}

\title{Quantitative Benchmarks for Theory Exploration}
  \author{Chris Warburton \\
          University of Dundee \\
          \texttt{cmwarburton@dundee.ac.uk}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
  \emph{Automated theory exploration} has the potential to help scientists,
  mathematicians and programmers discover unexpected properties of their models,
  theories and programs; but has so far been limited to small-scale experiments,
  with only a few, hand-picked, data points. We address this problem by
  re-purposing existing benchmarks from the theorem proving domain, producing an
  order of magnitude more data for evaluation, and use it to evaluate the
  performance of a state of the art exploration system. To combat the long run
  times of these systems, we also propose a pre-processing method to identify
  promising subsets of a theory, sacrificing completeness to gain speed. It is
  our hope that the availability of robust evaluation methods, and scalable
  algorithms, will encourage competition and innovation, making theory
  exploration more useful and relevant as a formal methods tool.
\end{abstract}

@@

Many fields represent knowledge using \emph{formal systems}, for example
scientific models, mathematical theories and computer programs. These systems
can be extended to meet new demands in two ways: the systems can be made more
complex, or many new systems can be defined. Complexification makes a system's
behaviour harder to predict, whilst proliferation dilutes the attention that can
be given to each particular system. This presents a challenge for experts
needing to characterise the behaviour of such systems, for example to prove
properties related to protocol security or mechanical safety.

One way to help these experts is through automation: formal systems are easily
represented as software, and techniques like \emph{automated theory exploration}
can help experts to discover properties of a system, whether expected or
surprising, desirable or not.

@@

An example of a simple formal system is the following software
library:

\begin{equation*}
  \begin{aligned}
    \name{append}(\name{nil},        z) &= z                                   \\
    \name{append}(\name{cons}(x,y),  z) &= \name{cons}(x, \name{append}(y, z)) \\
    \name{map}(f,           \name{nil}) &= \name{nil}                          \\
    \name{map}(f,    \name{cons}(x, y)) &= \name{cons}(f(x), \name{map}(f, y))
  \end{aligned}
\end{equation*}

These equations define two functions, $\name{append}$ (for joining one list to
another) and $\name{map}$ (for applying a function to the elements of a list),
based on the way they rearrange particular constants ($\name{nil}$ and
$\name{cons}$, which encode lists à la Lisp) and variables ($f$, $x$, $y$, $z$).
One \emph{property} of this library is that the following expressions are
equivalent:

\begin{equation}\label{eq:mapappend}
  \name{map}(f, \name{append}(x, y)) =
    \name{append}(\name{map}(f, x), \name{map}(f, y))
\end{equation}

This states that $\name{map}$ distributes over $\name{append}$, i.e.\ we can
perform $\name{map}$ and $\name{append}$ in either order and get the same
result. This is useful, since it may be much faster to run $\name{map}(f, x)$
and $\name{map}(f, y)$ in parallel on different machines, then $\name{append}$
the results; compared to $\name{append}$ing the inputs and running one big
$\name{map}$.\footnote{This is one half of the ``MapReduce'' model for parallel
  programming.~\cite{dean2008mapreduce}.}

Most automation of formal systems focuses on \emph{automated theorem proving}
(ATP), which dates back to the early days of artificial
intelligence~\cite{newell1956logic}. In ATP, the user gives a particular
property statement (like \ref{eq:mapappend}) to the program up front, and the
program searches for a proof. Modern ATP systems, such as
CVC4~\cite{barrett2011cvc4}, compete regularly to prove as many benchmark
problems as they can, as quickly as possible~\cite{sutcliffe2006state}.

\emph{Automated Theory Exploration} (ATE) is the task of taking in a formal
system (such as the definitions of $\name{map}$ and $\name{append}$ above) and
giving out interesting properties, like \ref{eq:mapappend}.\footnote{ATE doesn't
  require a proof of these properties; we can use ATP tools for that.} ATE has
seen little attention, partly because of its resource usage, and partly because
it is difficult to evaluate: what makes a property ``interesting''? Many general
measures have been proposed\cite{colton2000notion}, but for comparison purposes
we can stick to known theories and simply check if a human has bothered to state
these properties before!

Existing ATE programs have been evaluated this way on theories of lists and
natural numbers taken from the Isabelle proof
assistant~\cite{claessen2013automating}. The discovered properties are compared
against those in Isabelle, to see how many match up. Unfortunately these are the
only data sets used, and each is rather small (containing 9 and 12 properties,
respectively). This is not enough data from which to draw meaningful comparisons
or extrapolations.

\section{Contributions}

\subsection{Benchmarking}

We address the problem of ATE evaluation by implementing a much larger benchmark
suite, containing 343 properties of 269 definitions. We take these from an
existing ATP benchmark called TIP\cite{claessen2015tip}, since it
\begin{inlinelist}
\item separates definitions from theorems
\item allows higher-order functions (like $\name{map}$)
\item incorporates many tricky problems from the theorem-proving literature
\item contains problems of immediate relevance to software verification
\end{inlinelist}.

Each TIP property is given alongside all of its required definitions. We
re-purpose these for the ATE setting, by pooling together all of
the definitions\footnote{Including custom functions to construct and destruct
  constant values (like $\name{nil}$ and $\name{cons}$), we get 3595
  definitions, most of which are redundant copies used in many problems.
  Discarding these leaves 269 distinct definitions.} and all of the properties.

We provide a tool to sample subsets of these definitions, along with the subset
of properties which apply to that sample, so that theories of various sizes can
be generated for evaluating ATE systems.

\subsection{Scaling}\label{recurrent-clustering}

In the worst case, ATE scales exponentially with the number of definitions in a
theory. We have also investigated ways to mitigate this using \emph{relevance
  filtering}~\cite{meng2009lightweight}: given a large theory, can we explore
only those definitions which we expect to yield interesting properties? We have
implemented such a filter based on
\emph{recurrent clustering}~\cite{journals/corr/abs-1212-3618}, which compares
the syntax trees of a theory's definitions to find those which are similar. We
have also made a purely random filter, to control for the effect of
\emph{filtering itself}. These can be used as a pre-processor for any ATE system
and work is ongoing to evaluate their effect on property recall and performance.

\section{Results}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{runtimes}
  \caption{Median runtime of the QuickSpec ATE system for 30 iterations on
    theories of various size, sampled from TIP. Error bars show the
    interquartile range. We time out at 1000 seconds.}\label{fig:runtimes}
\end{figure}

Figure~\ref{fig:runtimes} shows the median run time of the existing
\textsc{QuickSpec}\cite{QuickSpec} system as we explore theories of various size
sampled from our benchmark. We have found that the time taken can vary greatly,
even among theories of the same size, with most finishing after a few seconds
but some taking hours; this can be seen by the increase in range for larger
theories, despite the median remaining low.

This variance makes current ATE systems inconvenient for widespread use. Finding
a pattern in these long-running inputs is current work; our hope is to make such
tools more predictable and tractable using approaches like our relevance filter.

\section{Future Work}\label{future-work}

There are many promising directions that this work enables. Our benchmark is
currently limited to ATE systems using Haskell; integrating other languages
would allow wider comparisons, e.g. to IsaCoSy and IsaScheme which use Isabelle.

Machine learning is a promising direction for scaling ATE, since its approximate
methods may be more efficient than existing searches, and its data-driven
approach is well suited to handle fuzzy concepts such as ``interestingness''.
Our benchmark can act as both a fitness function and a training corpus for such
efforts.

Our system can explore \emph{arbitrary} code, not just our benchmarks. Besides
listing properties, potential applications include tools for optimisation (if
coupled to a profiler) and refactoring (e.g. using properties to propose new
abstractions).

\bibliographystyle{plain}
\bibliography{./Bibtex}
